\MakeUppercase{Linear versus centred Colouring Numbers}\thanks{This research was partly funded by NSERC.}

### 2021-10-05 lin-vs-cen.tex
Specified some constants

\begin{lem}
  %For any $k\times k$ pseudogrid $H$
  Let $H$ be a $k\times k$ pseudogrid and $\mathcal{P}:=(P_x:x\in V(G_{k}))$ be such that
  \begin{inparaenum}[(i)]
      \item for each $x\in V(G_k)$, $H[P_x]$ is a path of length at most $r$ and
      \item $H/\mathcal{P}$ is a $\le\!\! r$-subdivision of $G_k$.
  \end{inparaenum}
  Then $\chilin(G)\in\Omega(k/r)$.
\end{lem}

\begin{proof}
  This proof uses two large integer constants $\alpha$ and $\beta$ and two small positive real constants $\epsilon$ and $\delta$.  The relationships between these constants are that $\beta = \Theta(\alpha^{1/4})$ and $\epsilon=\Theta(1/\alpha)$.  Let $\varphi:V(H)\to\{1,\ldots,C\}$ be an arbitrarily $C$-colouring of some $k\times k$ pseudogrid $H$ for some $C\le \epsilon k$.  We will show that $H$ contains a path $P$ which has no centre with respect to $\varphi$.

@@

  \begin{proof}
    This proof uses the notion of \emph{removing} a row or column of a pseudogrid.  Removing a row is defined as follows (removing a column is similar): For each vertex $v$ of $G^{\onesub}_k$ in the relevant row, remove all edges of $P_v$ from $H$. Next, remove any isolated vertices from $H$ and finally, repeatedly remove degree-$1$ vertices from $H$ until none remain.  The result is pseudogrid with one less row.

    While there exists some infrequent colour $c\in\{1,\ldots,C\}$, find a row or column of $G^{\onesub}_k$ that contains a vertex with $c$ in its colour set and remove that row or column.  Doing this exhaustively results in the removal of at most $(\alpha-1) C \le \epsilon\alpha k$ rows and at most $\epsilon\alpha k$ columns of $H$.  This leaves a pseudogrid $H'$ that contains a $k'\times k'$ pseudogrid for some $k'\ge (1-\epsilon \alpha)k$ for which every colour that appears in at least one vertex of $H'$ is frequent.
  \end{proof}

  Because of \cref{all_frequent}, we now assume that there is no infrequent colour, i.e., each colour $c\in\{1,\ldots,C\}$ has multiplicity at least $\alpha$.

@@

  \begin{enumerate}[{Stage} 1:]
    \item We begin this stage by defining each vertex $G^{\onesub}_k$ as being \emph{unmarked}.  Next we consider each $c\in\{1,\ldots,C\}$ in turn.  If $G^{\onesub}_k$ contains a pair $(x_c,y_c)$ where
    \begin{inparaenum}[(i)]
      \item each of $x_c$ and $y_c$ is unmarked;
      \item $\|x_cy_c\|_\infty \ge \beta$; and
      \item $c\in \phi(x_c)$ and $c\in \phi(y_c)$;
    \end{inparaenum}
    then we say that Stage~1 \emph{succeeeds} for colour $c$ and we mark each vertex in $B_{\beta}(x_c)$ and $B_\beta(y_c)$.

@@

    As mentioned above, for sufficiently large $\alpha$ (relative to $\beta$), the set $S$ is non-empty and therefore $R_1$ is non-empty.  Let $P:=R_1$ and let $Q:=\bigcup_{p\in P} B_\beta(p)$ and apply \cref{voronoi_mapping} to obtain the mapping $\vor:Q\to P$.  We use to this to construct a bipartite graph $I$ with vertex parts $\overline{S}$ and $R_1$. The edge set of $I$ contains an edge $cx$ precisely when there is a marked vertex $z$ of $G^{\onesub}_k$ such that $c\in\phi(z)$ and $\vor(z)=x$.

@@

    This implies that, for any $A_0\subseteq\overline{S}$,
    \[
        |N_I(A_0)| \ge \frac{a|A_0|}{r(7\beta/2+1)^2} \ge 2|N_I(A_0)| \enspace .
    \]
    for appropriately chosen $\alpha$ and $\beta$ (with $\beta\in O(\sqrt{\alpha}))$.\todo{Specify precise condition.}  Therefore, by \cref{hall_vees}, there exists a subgraph $\Lambda$ of $I$ in which each vertex of $A$ has degree $2$ and each vertex of $B$ has degree at most one.  Therefore, each $c\in A=\overline{S}$ has two $\Lambda$-neighbours $p$ and $q$ in $B$.  The edge $cp$ is present in $I$ because $G_k$ contains a vertex $x_c$ with $s(x_c)=p$ and $c\in \Phi(x_c)$.  Similarly, the edge $cq$ is present in $I$ because $G_k$ contains a vertex $y_c$ with $s(y_c)=q$ and $c\in\Phi(y_c)$.  In this way, the graph $\Lambda$ defines the pairs $\{x_c,y_c\}$ for each $c\in\overline{S}$.
  \end{enumerate}

@@

  Conditions~(\ref{never_far}) and (\ref{three_in_ball}) of \cref{voronoi_mapping} imply that each any ball of radius $\min\{\beta/8-1, \beta/12\}/2$ contains at most three elements of $R$.  Therefore, (for sufficiently large $\beta$) \cref{one_path_enchilada}, there exists a path $P$ in $G^{\onesub}_k$ that contains every vertex of $R$, that goes straight at every straight vertex of $R$ and that bends at every bent vertex of $R$. These latter two conditions guarantee that $V(P)$ contains $P_x$ for each $x\in R$. For each colour $c\in\{1,\ldots,C\}$, there are at least two vertices $x_c,y_c\in R$ such that $P_{x_c}$ and $P_{y_c}$ each contain some vertex of colour $c$.  Therefore the path $P$ contains two vertices of each colour.  Therefore $P$ has no center. Therefore $\varphi$ is not a linear centered colouring of $H$.  Since $\varphi$ is an arbitrary $\epsilon k$-colouring of $H$, this implies that $\chilin(H)\ge \epsilon k$.
\end{proof}

>>>>

\begin{lem}\label{pseudogrid_lower_bound}
  %For any $k\times k$ pseudogrid $H$
  Let $H$ be a $k\times k$ pseudogrid and $\mathcal{P}:=(P_x:x\in V(G_{k}))$ be such that
  \begin{inparaenum}[(i)]
      \item for each $x\in V(G_k)$, $H[P_x]$ is a path of length at most $r$ and
      \item $H/\mathcal{P}$ is a $\le\!\! r$-subdivision of $G_k$.
  \end{inparaenum}
  Then $\chilin(G)\in\Omega(k/r)$.
\end{lem}

\begin{proof}
  This proof uses two large integer constants $\alpha$ and $\beta$ and two small positive real constants $\epsilon$ and $\delta$.  The relationships between these constants are that $\alpha=\Theta(r\beta^2)$  $\epsilon=\Theta(1/\alpha)$.  Let $\varphi:V(H)\to\{1,\ldots,C\}$ be an arbitrarily $C$-colouring of some $k\times k$ pseudogrid $H$ for some $C\le \epsilon k$.  We will show that $H$ contains a path $P$ which has no centre with respect to $\varphi$.

@@

  \begin{proof}[Proof of \cref{all_frequent}]
    This proof uses the notion of \emph{removing} a row or column of a pseudogrid.  Removing a row is defined as follows (removing a column is similar): For each vertex $v$ of $G^{\onesub}_k$ in the relevant row, remove all edges of $P_v$ from $H$. Next, remove any isolated vertices from $H$ and finally, repeatedly remove degree-$1$ vertices from $H$ until none remain.  The result is pseudogrid with one less row.

    While there exists some infrequent colour $c\in\{1,\ldots,C\}$, find a row or column of $G^{\onesub}_k$ that contains a vertex with $c$ in its colour set and remove that row or column.  Doing this exhaustively results in the removal of at most $(\alpha-1) C \le \epsilon\alpha k$ rows and at most $\epsilon\alpha k$ columns of $H$.  This leaves a pseudogrid $H'$ that contains a $k'\times k'$ pseudogrid for some $k'\ge (1-\epsilon \alpha)k$ for which every colour that appears in at least one vertex of $H'$ is frequent.  This completes the proof of \cref{all_frequent}.
  \end{proof}

  Returning to the proof of \cref{pseudogrid_lower_bound}, we set $\epsilon=1/2\alpha=O(1/r\beta^2)$ so that \cref{all_frequent} guarantees that $H'$ contains a $k/2\times k/2$ pseudogrid in which each colour $c\in\{1,\ldots,C\}$ either does not appear in $H'$ or is frequent in $H'$.  To avoid overloading notation, we now assume that there is no infrequent colour, i.e., each colour $c\in\{1,\ldots,C\}$ has multiplicity at least $\alpha$ in $H$.  (The remainder of the proof does not depend on the values of $k$ or $C$.)

@@

  \begin{enumerate}[{Stage} 1:]
    \item We begin this stage by defining each vertex $G^{\onesub}_k$ as being \emph{unmarked}.  Next we consider each $c\in\{1,\ldots,C\}$ in turn.  If $G^{\onesub}_k$ contains a pair $(x_c,y_c)$ where
    \begin{inparaenum}[(i)]
      \item each of $x_c$ and $y_c$ is unmarked;
      \item $\|x_cy_c\|_\infty \ge \beta$; and
      \item $c\in \Phi(x_c)$ and $c\in \Phi(y_c)$;
    \end{inparaenum}
    then we say that Stage~1 \emph{succeeeds} for colour $c$ and we mark each vertex in $B_{\beta}(x_c)$ and $B_\beta(y_c)$.

@@

    As mentioned above, for sufficiently large $\alpha$ (relative to $\beta$), the set $S$ is non-empty and therefore $R_1$ is non-empty.  Let $P:=R_1$ and let $Q:=\bigcup_{p\in P} B_\beta(p)$ and apply \cref{voronoi_mapping} to obtain the mapping $\vor:Q\to P$.  We use to this to construct a bipartite graph $I$ with vertex parts $\overline{S}$ and $R_1$. The edge set of $I$ contains an edge $cx$ precisely when there is a marked vertex $z$ of $G^{\onesub}_k$ such that $c\in\Phi(z)$ and $\vor(z)=x$.

@@

    This implies that, for any $A_0\subseteq\overline{S}$,
    \[
        |N_I(A_0)| \ge \frac{\alpha|A_0|}{r(7\beta/2+1)^2} \ge 2|A_0| \enspace .
    \]
    for $\alpha\ge 2r(7\beta/2+1)^2$.  Therefore, by \cref{hall_vees}, there exists a subgraph $\Lambda$ of $I$ in which each vertex of $A$ has degree $2$ and each vertex of $B$ has degree at most one.  Therefore, each $c\in A=\overline{S}$ has two $\Lambda$-neighbours $p,q\in R_1$.  The edge $cp$ is present in $I$ because $G_k$ contains a vertex $x_c$ with $\vor(x_c)=p$ and $c\in \Phi(x_c)$.  Similarly, the edge $cq$ is present in $I$ because $G_k$ contains a vertex $y_c$ with $\vor(y_c)=q$ and $c\in\Phi(y_c)$.  In this way, the graph $\Lambda$ defines the pairs $\{x_c,y_c\}$ for each $c\in\overline{S}$.
  \end{enumerate}

@@

  Conditions~(\ref{never_far}) and (\ref{three_in_ball}) of \cref{voronoi_mapping} imply that each any ball of radius $\min\{\beta/8-1, \beta/12\}/2$ contains at most three elements of $R$.  This condition is satisfied for $\beta= 12\cdot18 = 218$ (so the conditions on $\alpha$ are satisified for any $\alpha = r(7\beta/2+1)^2=2331729r$).  Therefore, by \cref{one_path_enchilada}, there exists a path $P$ in $G^{\onesub}_k$ that contains every vertex of $R$, that goes straight at every straight vertex of $R$ and that bends at every bent vertex of $R$. These latter two conditions guarantee that $V(P)$ contains $P_x$ for each $x\in R$. For each colour $c\in\{1,\ldots,C\}$, there are at least two vertices $x_c,y_c\in R$ such that $P_{x_c}$ and $P_{y_c}$ each contain some vertex of colour $c$.  Therefore the path $P$ contains two vertices of each colour, so $P$ has no center and therefore $\varphi$ is not a linear colouring of $H$.  Since $\varphi$ is an arbitrary $\epsilon k$-colouring of $H$, this implies that $\chilin(H)\ge \epsilon k = \Theta(k/\alpha)=\Theta(k/r)$.
\end{proof}
